Evolution of \\ ice sheet geometry \\ using Stokes dynamics

### 2021-06-21 talk/talk.tex
Edit talk.

\normalsize
\vspace{10mm}
\begin{itemize}
\item<2> work in progress, with no results yet
    \begin{itemize}
    \item<2> low expectations, please
    \end{itemize}
\item<2> I am giving this talk at 4am my time
    \begin{itemize}
    \item<2> ditto
    \end{itemize}
\end{itemize}
\end{frame}

@@

\vspace{-1mm}
\begin{itemize}
\item glacier geometry which evolves according to climate
    \begin{itemize}
    \item simplest such problem (grounded, nonsliding, isothermal)
    \end{itemize}
\item assume time-independent data on fixed $\Omega \subset \RR^2$:
    \begin{itemize}
    \item (net) climatic mass balance $a$ \hfill $\gets$ \emph{$a>0$ for accumulation}
    \item bed elevation $b$
    \end{itemize}
\item \emph{to compute}: surface elevation $s$ on $\Omega$ and icy domain $\Lambda_s \subset \RR^3$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item glaciers flow
\item $\therefore$ must solve an equation for the surface elevation
\item the surface kinematical equation (SKE) holds \emph{on} the ice
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 & &\text{\emph{evolving}} \\
-\bu|_s \cdot \bn_s - a &= 0 & &\text{\emph{steady state}}
\end{align*}

@@

\begin{itemize}
\item the steady IGP free-boundary problem, for $s(x,y)$ on $\Omega$, is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a non-Newtonian Stokes problem for $\bu,p$:}

@@

\begin{frame}{\alert{implicit} IGP strong form}

@@

\begin{itemize}
\item almost no one is solving the implicit IGP
    \begin{itemize}
    \item except Bueler (2016), Brinkerhoff et al (2017) (\emph{shallow})
    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, fake ice layer})
    \item none are scalable (\emph{single level})
    \end{itemize}
\item what are people doing instead, to get science?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
    \item often with shallow approximations, e.g.~PISM run below

\medskip
\begin{columns}
    \column{0.7\textwidth}
        \includegraphics[width=0.8\textwidth]{figs/seguinot.png}
    \column{0.3\textwidth}
        \scriptsize \emph{by Julien Seguinot}
\end{columns}

\medskip
    \item \emph{my goals}: Stokes, unconditionally-stable steps, scalable
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0,$$
extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
\item regarding Stokes problem for fixed geometry:
    \begin{itemize}
    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz 2011)
    \item optimal solver exists (Isaac et al 2015)
    \end{itemize}
\item $\therefore$ \, $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
    \begin{itemize}
    \item which is a regularity conjecture for the coupled theory
    \end{itemize}
\end{itemize}

@@

\item well-posedness is completely open
    \begin{itemize}
    \item but existence holds for SIA version (Jouvet \& Bueler, 2012)
    \end{itemize}

@@

    \begin{itemize}
    \item $\Phi$ is elliptic in the large
    \item $\Phi$ is degenerate at the ice margin
    \end{itemize}
\item yet common to say
    \begin{itemize}
    \item $\Phi$ is advective because SKE is write-able as a transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
    \end{itemize}
\end{itemize}
\end{frame}

>>>>

\normalsize
\vspace{10mm}
\begin{itemize}
\item<2-3> work in progress
\item<2-3> I am giving this talk at 4am my time
    \begin{itemize}
    \item<3> low expectations, please
    \end{itemize}
\end{itemize}
\end{frame}

@@

\vspace{-1mm}
\begin{itemize}
\item glacier geometry evolution according to the climate?
    \begin{itemize}
    \item simplest such problem (grounded, nonsliding, isothermal)
    \end{itemize}
\item assume time-independent data on fixed $\Omega \subset \RR^2$:
    \begin{itemize}
    \item (net) climatic mass balance $a(x,y)$ \hfill $\gets$ \emph{$a>0$ for accumulation}
    \item bed elevation $b(x,y)$
    \end{itemize}
\item \emph{to compute}: surface elevation $s(t,x,y)$ on $\Omega$ \emph{and} icy domain $\Lambda_s \subset \RR^3$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item glaciers flow, so one must solve the surface kinematical equation (SKE) \emph{on} the ice
    $$\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a = 0$$

@@

\begin{itemize}
\item the \alert{steady} IGP is a \alert{nonlinear complementarity problem (NCP) for $s$}, a free-boundary problem, \alert{which is coupled to a non-Newtonian Stokes problem for $\bu,p$:}

@@

\begin{frame}{implicit IGP strong form}

@@

\begin{itemize}
\item almost no one is solving such a implicit or steady IGP
    \begin{itemize}
    \item except Bueler (2016), Brinkerhoff et al (2017) (\emph{shallow})
    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, fake ice layer})
    \item none are scalable (\emph{single level})
    \end{itemize}
\item what are people doing instead, for production science?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
    \item often with shallow approximations, e.g.~PISM run below
    \end{itemize}

\medskip
\begin{columns}
    \column{0.7\textwidth}
        \includegraphics[width=0.8\textwidth]{figs/seguinot.png}
    \column{0.3\textwidth}
        \scriptsize \emph{by Julien Seguinot}
\end{columns}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0,$$
extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
\item regarding this Stokes problem (for \emph{fixed} geometry):
    \begin{itemize}
    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz 2011)
    \item optimal solver exists (Isaac et al 2015)
    \end{itemize}
\item $\therefore$ \, $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
    \begin{itemize}
    \item which is a regularity conjecture
    \end{itemize}
\end{itemize}

@@

\item well-posedness is almost completely open
    \begin{itemize}
    \item existence holds for SIA version (Jouvet \& Bueler, 2012)
    \end{itemize}

@@

    \begin{itemize}
    \item $\Phi$ is elliptic ``in the large''
    \item $\Phi$ is degenerate at the ice margin
    \end{itemize}
\item yet common to say
    \begin{itemize}
    \item $\Phi$ is advective because SKE is write-able as a transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
    \end{itemize}
\end{itemize}
\end{frame}
